---
title: "Analysing the U.S. National Oceanic and Atmospheric Administration’s storm database to find the severe weather events"
author: "S. Meiyalaghan"
date: "`r Sys.Date()`"
output: 
    github_document:
        toc: yes
---


## Synopsis
The U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database was explored to find the severe weather events. This data analysis has been performed in partial fulfillment of the requirements for the [Reproducible Research](https://www.coursera.org/learn/reproducible-research) course available online from Johns Hopkins University via Coursera. In this analysis, dataset from the year 1995 was taken into account following the visualisation of data distribution (Figure). The fifteen most harmful weather events with respect to population health across the United States have been identified, mainly based on the number of fatalities, and listed in Table 1. Furthermore, the fifteen weather events that have greatest economic consequences across the United States have been identified, mainly based on the property damaged, and listed in Table 2.



## Data Processing

#### Key processes:

* Load required libraries to perform this analysis

* Download the storm data
        The details of constructed/defined variables in the storm data were obtained from the following documentation
        
1. [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdffiles)

2. [National Climatic Data Center Storm Events FAQ](https://www.ncdc.noaa.gov/stormevents/faq.jsp)
          
          
* Read and inspect the downloaded data

* Convert the variable to the right format 

* Create a new variable called 'Year' 

* Visualise the storm data distribution
 


**Load required libraries**
```{r}
#load require libraries
library(data.table) # required for 'fread'
library(magrittr) #required to use the pipe (%>%) operator
library(dplyr) # for efficiently manipulating datasets
library(pander) # to produce good tables
```




#### Download the file
```{r}
# create a local directory named "data" if not exsits to download the file 
localDir <- "data"
if (!file.exists(localDir)) {
  dir.create(localDir)
}

# download the file if not exsits
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
stormData_file <- paste(localDir, basename(url), sep='/')
if (!file.exists(stormData_file)) {
  download.file(url, stormData_file)
}

```



#### Read and inspect the data
```{r}
#read the data
stormData <- read.csv(stormData_file, as.is = TRUE)

# A freshly read data.frame should always be inspected with functions such as head, str, summary and tail

head(stormData)
str(stormData)
summary(stormData)

```


#### Converting date variable to date format and create a new variable for years
```{r}
#convert date variable named 'BGN_DATE'  to `Date` format, then extract and list the years in a  newly created variable called 'Year'.
stormData$Year <- year(as.Date(stormData$BGN_DATE, '%m/%d/%Y'))

range(stormData$Year) # to see the start and end of data year, 
```

####  Visualisation of storm data distribution
```{r}
# Plotting a histogram for a quick visualisation of strom data distribution by year
#It is important to select the correct 'bin' size (groups of data); here 61 is applied since range of Year from 1950 to 2011 
hist(stormData$Year, 61)
```

*This histogram plots shows a "skewed left" (non-symmetric) distribution in which the tail is on the left side. This could be due to the lack of data recorded in the early years. Given this "skewed left" distribution, data from 1995 are taken for this analysis.*


## Results

#### Tyeps of weather events across the United States that are most harmful with respect to population health 


```{r}
most_harmful_events <- stormData %>%
                    subset(Year >= "1995") %>% #subset dataset with  from 1995
                    select(EVTYPE, FATALITIES, INJURIES) %>%
                    group_by(EVTYPE) %>% 
                    summarise_all(funs(sum)) %>%
                    arrange(desc(FATALITIES)) %>% 
                    rename("TYPE OF EVENT" = EVTYPE)

```



```{r}
#create a table to list the 15 most harmful weather events
pandoc.table((most_harmful_events[1:15,]),  
          justify = "left",
          style = "grid",
          caption = "1. The fifteen most harmful weather events with respect to population health across the United States")

```

*Table 1 shows the fifteen harmful weather events, arranged mainly by the number of fatalities, from the year 1995.*


#### Tyeps of weather events across the United States that have greatest economic consequences 

```{r}
#finding greatest economic consequences events from the storm dataset
greatest_economic_events <- stormData %>%
                    subset(Year >= "1995") %>% #subset dataset with  strom data from 1995
                    select(EVTYPE, PROPDMG, CROPDMG) %>%
                    group_by(EVTYPE) %>% 
                    summarise_all(funs(sum)) %>%
                    arrange(desc(PROPDMG), desc(CROPDMG)) %>% 
                    rename("TYPE OF EVENT" = EVTYPE,
                           "PROPERTY DAMAGED" = PROPDMG,
                           "CROP DAMAGED" = CROPDMG)
```



```{r}
#create a table to list  the 15 greatest economic consequences events
pandoc.table((greatest_economic_events[1:15,]),
            justify = "left",
            style = "grid",
            caption = "2. The fifteen weather events that have greatest economic consequences across the United States")
```

*Table 2 shows the fifteen weather events that have greatest economic consequences, arranged mainly by the cost of property damaged, from the year 1995.*


